---
title: 'CSCI E-63C: Final Exam/Project'
output:
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(splus2R)
library(e1071)
library(MASS)
library(randomForest)
# library(ggplot2)
# library(GGally)
```

# Preface

For the final exam/project we will develop classification models using several approaches and compare their performance on a new dataset that is a subset of one of the datasets used in machine learning common task framework (CTF) competitions.  A copy of it split into training (`final-data-train.csv`, with outcome `response` available) and test (`final-data-test.csv`, stripped of the outcome, for prediction purposes only) datasets is available on our course website in Canvas as a zip-archive of all associated files.

Please notice, that at the end of this final exam/project you will be asked, in addition to the Rmarkdown and HTML files, to also make predictions for the observations in the *test* (not training!) dataset and upload them into Canvas as well.  The expected format for the file with predictions for test dataset is two columns of comma-separated values, one row per observation in *test* dataset, first column -- the observation identifier (column `id` in test dataset) and the second column -- your best model predictions for each observation in the *test* dataset as Y/N indicator.  To illustrate expected format the zip archive contains also a couple of examples of test predictions in this format for your reference as well (`predictions-*.csv` files in `predictions-examples` sub-folder in zip-archive).

One more time, to iterate and emphasize, please notice that this time your submission must consist of the following *three* (not just two, Rmd+html, as usual) items:

* Rmarkdown *.Rmd file with all the calculations you want to receive credit for,
* HTML version of the output generated by your *.Rmd file, and
* **predictions** for the **test** dataset in comma-separated values (CSV) format (file name **must** have *.csv extension for the file to load in Canvas)

The teaching team invites you to load your predictions (just predictions, just for the test dataset according to the file format shown in the sample files in the zip-archive) into Canvas repeatedly as you work on your models and improve them over the course of this week.  At least daily (or more frequently as we see fit) we will download predictions loaded by everyone by that time and compile a leaderboard in html format of all of them sorted by their accuracy as compared to the true values of the outcome for the test dataset (along with their sensitivity, specificity, etc.).  This list will be made available on our course website in Canvas for everyone in this class in order to see how the performance of their models compares across the rest of the models built by other students in the class.  The first version of the leaderboard posted on the course website at the time when final exam is made available starts with predictions made by those few example files provided in the zip-archive (coin flip, majority vote, etc. -- what do you think Charlie Brown is using to make the predictions?).  Those should be pretty easy to improve upon.

It is 100% up to you whether you want to upload your model predictions over the course of this week, how frequently you want to do it and what you want its results to be called in the leaderboard posted for everyone in the class to see.  We will use the name of the 2nd column in the file with predictions (the one containing Y/N values, not the numerical ids of the observations) as the model name listed in the leaderboard.  If you prefer not to use your name, choose something else instead, sufficiently unique so that it is easier for you to spot your result among all others.  Once again, please check out sample files of dull (majority vote, coin flip, etc.) predictions we have made available and consider how they show up in the leaderboard html file already posted on Canvas website.  Once you are done with final you are expected to load predictions from your best model into Canvas -- that is part of your points total as explained below.

Lastly, the size of this dataset can make some of the modeling techniques run slower than what we were typically encountering in this class.  You may find it helpful to do some of the exploration and model tuning on multiple random samples of smaller size as you decide on useful ranges of parameters/modeling choices, and then only perform a final run of fully debugged and working code on the full dataset.  Please see also the afterword below on the computational demands of this problem set.

# Problem 1: univariate and unsupervised analysis (20 points)

Download and read training and test data into R and prepare graphical and numerical summaries of it: e.g. histograms of continuous attributes, contingency tables of categorical variables, scatterplots of continuous attributes with some of the categorical variables indicated by color/symbol shape, etc.  Whatever you find helpful to think about properties of the data you are about to start using for fitting classification models.

As it is often the case for such contests, the attributes in the dataset are blinded in the sense that no information is available about what those are or what their values mean.  The only information available is that the attribute `response` is the outcome to be modeled and the attribute `id` is the unique numerical identifier for each observation.  Some of the remaining attributes are clearly categorical (those that are character valued) and some rather obviously continuous (those with numerical values with large number of unique values).  For several of them it is less clear whether it is best to treat them as continuous or categorical -- e.g. their values are numerical but there are relatively few unique values with many observations taking the same value, so that they arguably could be treated as continuous or categorical.  Please idenify them, reflect on how you prefer to handle them and describe this in your own words.

Perform principal components analysis of this data (do you need to scale it prior to that? how would you represent multilevel categorical attributes to be used as inputs for PCA?) and plot observations in the space of the first few principal components indicating levels of some of the categorical attributes of your choosing by the color/shape of the symbol.  Perform univariate assessment of associations between the outcome we will be modeling and each of the attributes (e.g. t-test or logistic regression for continuous attributes, contingency tables/Fisher exact test/$\chi^2$ test for categorical attributes).  Summarize your observations from these assessments: does it appear that there are predictors associated with the outcome `response` univariately? Which predictors seem to be more/less relevant?

``` {r problem1pt1}
trainData <- read.table("final-data-train.csv", sep=",", header=TRUE)
testData <- read.table("final-data-test.csv", sep=",", header=TRUE)

head(trainData)
summary(trainData)
str(trainData)
old.par <- par(mfrow=c(2,2), ps=16)
for (col in 2:ncol(trainData)) {
  if (class(trainData[,col]) != "factor") {
    hist(trainData[,col], main=paste("Hist of", colnames(trainData[col])))
  }
}

old.par <- par(mfrow=c(2,2), ps=16)
hist(trainData$dtj)
scaledDtj <- scale(trainData$dtj)
hist(scaledDtj)

hist(trainData$is)
scaledIs <- scale(trainData$is)
hist(scaledIs)

hist(trainData$bw)
scaledBw <- scale(trainData$bw)
hist(scaledBw)

trainData$is <- scaledIs
```

**To begin, I generated a simple numerical summary as well as examined the R-given data types for each variable. Noticing that the variable "ypz" was fixed for integer values between 0 and 17, I considered interpreting this as a factor. However, because observations take on all values in the range (17 total levels), this was above the R-suggested max number of levels of 15, so I will interpret as a continuous variable. I then generated histograms of each continuous variable to get an idea of their distributions. For continuous variables with a strong tail in distribution, I tried scaling to see if it would result in a more symmetric distribution. I found significant improvement only in the variable "is", so I scaled it in the original dataset.**

``` {r problem1pt2}

old.par <- par(mfrow=c(2,2), ps=16)
for (i in 2:ncol(trainData)) {
  if (class(trainData[,i]) != "factor") {
    plot(trainData[,i], col=trainData$response, ylab=colnames(trainData[i]), main=paste(colnames(trainData[i]), "colored by Response"))
    plot(trainData[,i], col=trainData$wc, ylab=colnames(trainData[i]), main=paste(colnames(trainData[i]), "colored by WC"))
    plot(trainData[,i], col=trainData$wi, ylab=colnames(trainData[i]), main=paste(colnames(trainData[i]), "colored by WI"))
    plot(trainData[,i], col=trainData$ent, ylab=colnames(trainData[i]), main=paste(colnames(trainData[i]), "colored by ENT"))
  }
}


plot(trainData$bw, col=trainData$response)

for (i in seq(2,ncol(trainData),2)) {
  if (class(trainData[,i]) == "factor") {
    for (j in seq(2,ncol(trainData),2)) {
      if (class(trainData[,j]) == "factor") {
        if (i != j) {
          print(table(trainData[,i], trainData[,j]))
        }
      }
    }
  }
}
```

**Next, I generated single-variable plots colored by 4 different factor variables to see if there were any easily-identifiable patterns in the data set. One clear pattern is a distinction in response in the continuous variable "qh". It certainly makes it more difficult not knowing the meaning of these variables but such a pattern with the outcome variable for classification leads me to believe that it will be an important predictor. Finally, I generated some contingency tables of categorical variables to hopefully scope out some covariance. It is difficult to discern patterns from the data as any apparent patterns may be a result of the distribution of a given categorical variable rather than covariance between two.**

``` {r problem1pt3}
trainDataDummy <- trainData
# Get column indices or names of factor vars from dataset
factors <- Filter(is.factor, trainData)

for (i in 1:ncol(factors)) {
  for (level in unique(factors[,i])) {
    trainDataDummy[paste("dummy", colnames(factors[i]), level, sep = "_")] <- ifelse(trainDataDummy[,i] == level, 1, 0)
  }
}

trainDataDummy <- trainDataDummy[,-c(1:5, 7, 9, 13:18)]

pca <- prcomp(trainDataDummy)
plot(pca)
biplot(pca)

pvalues <- vector()

for (i in 1:ncol(trainData)) {
  if (is.factor(trainData[,i])) {
    table <- table(trainData$response, trainData[,i])
    chisqtest <- chisq.test(table)
    pvalues[i] <- chisqtest$p.value
  }
  else {
    logreg <- glm(response~trainData[,i],data=trainData, family=binomial)
    pvalues[i] <- coef(summary(logreg))[4]
  }
}

pvalues <- as.data.frame(pvalues)
rownames(pvalues) <- colnames(trainData)
pvalues
```

**To run a PCA, I generated dummy variables to represent each level of each categorical variable in the dataset. The PCA does not appear to be significantly driven by one single variable, so we may have several contributing significantly to prediction for models ahead. Next, I iterated through the dataset and generated univariate models - chi squared models for categorical variables and logistic regression for continuous variables. Storing each p-value, I was able to compare relevance of each variable to each other. It appears that some categorical variables such as "tdt" and "np" may be predictive in a univariate setting. This is not a test from which conclusions should be drawn, moreso an investigation + preparation for including these variables in multivariate tests.**

# Problem 2: logistic regression (20 points)

Develop logistic regression model of the outcome `response` as a function of multiple predictors in the model.  Which variables are significantly associated with the outcome?  Test model performance on multiple splits of data into training and test subsets and summarize it in terms of accuracy/error/sensitivity/specificity.

``` {r problem2pt1}
trainDataNoID <- trainData[,-2]
head(trainDataNoID)

logreg <- glm(response~.,data=trainDataNoID, family=binomial)
summary(logreg)
```

**Up until this point, there were no consequences from including the "id" variable in the dataset, as there was no prediction or multivariate modeling occurring. Now, we do not want to include a variable that we know to not have statistical / predictive meaning, so I've subset the data to no longer include id. I ran a single logistic regression model on the entire training set just to get an idea of significant predictor variables. From this model, it appears that the categorical variable "zwp" is significant, as almost all dummy variables generated for the categorical variable are deemed significant. Categorical variables "ku" and "sb" as well as continuous variables "is", "npk", and "qh" may be of similar importance.**

``` {r problem2pt2}

accuracy <- vector()
specificity <- vector()
sensitivity <- vector()

for (iTry in 1:10) {
  trainIdx <- sample(nrow(trainDataNoID),nrow(trainDataNoID), replace=TRUE)
  wTrain <- trainDataNoID[trainIdx,]
  wTest <- trainDataNoID[-trainIdx,]
  logreg <- glm(response~.,data=trainDataNoID, family=binomial)
  glmTestRes <- as.numeric(predict(logreg,newdata=wTest,type="response")>0.5)
  tblTstglm <- table(wTest[,"response"],glmTestRes)

  accuracy[iTry] <- sum(diag(tblTstglm))/sum(tblTstglm)
  specificity[iTry] <- tblTstglm[1,1] / sum(tblTstglm[,1])
  sensitivity[iTry] <- tblTstglm[2,2] / sum(tblTstglm[,2])
}

old.par <- par(mfrow=c(1,3))
boxplot(accuracy, col="red", main="Accuracy")
boxplot(specificity, col="blue", main="Specificity")
boxplot(sensitivity, col="green", main="Sensitivity")
```

**I ran 10 iterations repeatedly splitting the data into training and testing, then running a logistic regression model to predict response as a function of all other variables. The results show a median accuracy of around 0.887 across the 10 trials. The specificity is quite similar, falling at about 0.898. However, the sensitivity is only about 0.839 across the iterations. This means that the logistic regression model is missing a significant number of positive cases, or otherwise put, the true positive rate is lower than the true negative rate.**

## Extra points problem: interaction terms (5 extra points)

Assess the impact/significance of pairwise interaction terms for all pairwise combinations of covariates used in the model and report the top ten that most significantly improve model fit.

# Problem 3: linear discriminant analysis (15 points)

Fit linear discriminant analysis model of the outcome `response` as a function of the rest of covariates in the dataset.  Feel free to decide whether you want to use all of them or a subset of those.  Test resulting model performance on multiple splits of the data into training and test subsets, summarize it in terms of accuracy/error/sensitivity/specificity and compare them to those obtained for logistic regression.

``` {r problem3pt1}
accuracy <- vector()
specificity <- vector()
sensitivity <- vector()

for (iTry in 1:10) {
  trainIdx <- sample(nrow(trainDataNoID),nrow(trainDataNoID), replace=TRUE)
  wTrain <- trainDataNoID[trainIdx,]
  wTest <- trainDataNoID[-trainIdx,]
  ldaTry <- lda(response~.,data=wTrain)
  ldaTestRes <- predict(ldaTry,newdata=wTest)$class
  tblTstLDA <- table(wTest[,"response"],ldaTestRes)

  accuracy[iTry] <- sum(diag(tblTstLDA))/sum(tblTstLDA)
  specificity[iTry] <- tblTstLDA[1,1] / sum(tblTstLDA[,1])
  sensitivity[iTry] <- tblTstLDA[2,2] / sum(tblTstLDA[,2])
}

old.par <- par(mfrow=c(1,3))
boxplot(accuracy, col="red", main="Accuracy")
boxplot(specificity, col="blue", main="Specificity")
boxplot(sensitivity, col="green", main="Sensitivity")
```

**I followed a very similar procedure for an LDA model. When compared to the logistic regression model from above, we see a median rate of accuracy here that is just lower, by about 1%. Specificity follows suit, with a median rate at about 1% lower than logistic regression once again. The sensitivity rate is actually higher with LDA, but with a lower overall accuracy rate this may not be convincing enough to consider it a better fit model for the data.**

# Problem 4: random forest (15 points)

Develop random forest model of outcome `response`. Present variable importance plots and comment on relative importance of different attributes in the model.  Did attributes showing up as more important in random forest model also appear as significantly associated with the outcome by logistic regression?  Test model performance on multiple splits of data into training and test subsets, compare test and out-of-bag error estimates, summarize model performance in terms of accuracy/error/sensitivity/specificity and compare to the performance of logistic regression and LDA models above.

``` {r problem4pt1}

accuracy <- vector()
specificity <- vector()
sensitivity <- vector()

old.par <- par(mfrow=c(1,2))

for (iTry in 1:10) {
  trainIdx <- sample(nrow(trainDataNoID),nrow(trainDataNoID), replace=TRUE)
  wTrain <- trainDataNoID[trainIdx,]
  wTest <- trainDataNoID[-trainIdx,]
  
  rf <- randomForest(wTrain[,-1], wTrain$response)
  
  if (iTry %% 3 == 0) {
    varImpPlot(rf)
  }

  rfpred <- predict(rf, newdata=wTest[,-1], type="class")
  confmatrix <- table(rfpred, wTest$response)
  
  accuracy[iTry] <- sum(diag(confmatrix))/sum(confmatrix)
  specificity[iTry] <- confmatrix[1,1] / sum(confmatrix[,1])
  sensitivity[iTry] <- confmatrix[2,2] / sum(confmatrix[,2])
}

old.par <- par(mfrow=c(1,3))
boxplot(accuracy, col="red", main="Accuracy")
boxplot(specificity, col="blue", main="Specificity")
boxplot(sensitivity, col="green", main="Sensitivity")

```

**The results of the random forest show a notch better results, with a median accuracy of about 0.897. Here there is a larger disparity between specificity and sensitivity, with specificity falling at about 0.942 and sensitivity at only 0.76. The fact that it's missing almost a quarter of positive cases makes me hesitant to use it as the predicting model on an unseen data set.**

# Problem 5: SVM (20 points)

Develop SVM model of categorical outcome `response` deciding on the choice of kernel, cost, etc. that appear to yield better performance.  Test model performance on multiple splits of data into training and test subsets, summarize model performance in terms of accuracy/error/sensitivity/specificity and compare to the performance of the rest of the models developed above (logistic regression, LDA, random forest).

``` {r problem5pt1}

accuracy <- vector()
specificity <- vector()
sensitivity <- vector()

iTry <- 1
for (iTry in 1:10) {
  trainIdx <- sample(nrow(trainDataNoID), nrow(trainDataNoID)/100, replace=TRUE)
  wTrain <- trainDataNoID[trainIdx,]
  wTest <- trainDataNoID[-trainIdx,]
  tune.out=tune(svm,response~.,data=wTrain, ranges=list(cost=c(0.1, 0.5, 0.97, 1, 1.05, 1.25), gamma=c(0, 0.01, 0.05, 0.1, 0.2, 0.3),  kernel=c("linear", "radial")))

  bestmod <- tune.out$best.model
  ypred <- predict(bestmod, wTest)
  resultstable <- table(predict=ypred, truth=wTest$response)

  accuracy[iTry] <- sum(diag(resultstable))/sum(resultstable)
  specificity[iTry] <- resultstable[1,1] / sum(resultstable[,1])
  sensitivity[iTry] <- resultstable[2,2] / sum(resultstable[,2])
}

old.par <- par(mfrow=c(1,3))
boxplot(accuracy, col="red", main="Accuracy")
boxplot(specificity, col="blue", main="Specificity")
boxplot(sensitivity, col="green", main="Sensitivity")

```

**The output of the SVM model unfortunately makes worse the trend that I've seen in the last few models. We have a median accuracy that's almost on par with the rest (~0.86) and a specificity that's even higher (~0.953), but a huge loss in sensitivity, with a median around (0.57). Although the model accuracy % is in the high 80s, we are still missing almost half of the true cases. This model, tuned to optimize 3 different parameters, most likely cannot outperform the above models on the basis of such a low sensitivity.**

# Problem 6: predictions for test dataset  (10 points)

## Problem 6a: compare logistic regression, LDA, random forest and SVM model performance (3 points)

Compare performance of the models developed above (logistic regression, LDA, random forest, SVM) in terms of their accuracy, error and sensitivity/specificity.  Comment on differences and similarities between them.

**The models have some interesting consistencies across types but also differ significantly in some quantitative aspects. All models are similar in having higher specificity rates than sensitivity. For the given dataset, these classification models are much more successful in correctly predicting negative outcomes than positive ones. Each model has struggled with sensitivity to some degree. Another point I was looking out for was variance in accuracy rates. Even if a median accuracy rate was higher in one model than another, I would like to see a narrower distribution across the 10 iterations to ensure that similar results would translate in the test data set. I was pleased to find fairly consistent distributions in accuracies, with IQR's at about 0.002 to 0.003 across all models. This allowed me to more confidently focus on accuracy when choosing on optimal model.**

## Problem 6b: make predictions for the **test** dataset (3 points)

Decide on the model that performs the best and use it to make predictions for the **test** dataset.  This is the dataset that is provided separately from training data without the outcome `response` that we are modeling here.  Upload resulting predictions in comma-separated values (CSV) format into the Canvas website.  Please check sample files with test dataset predictions for the expected format of the *.csv file with predictions: your submission must be in precisely the same format -- two and only two columns, first column - ids of the test observations ("id" column in test dataset), second - predictions as Y/N calls (not 0/1, 1/2, true/false, etc.).  The name of the second column of predictions is what will be used in leaderboard as its name.

``` {r problem6pt1}

testDataNoID <- testData[,-1]
head(testDataNoID)

rfFinalPred <- predict(rf, newdata=testDataNoID, type="class")
rfFinalPred <- as.data.frame(rfFinalPred)
rfFinalPred["id"] <- testData[,"id"]
rfFinalPred["DevanV"] <- rfFinalPred[,"rfFinalPred"]
myPredictions <- rfFinalPred[,-1]
write.csv(myPredictions, "devan-venkataraman-predictions.csv")
```

## Problem 6c: get better than coin flip by 10% (4 points)

This is not really a problem *per se* but rather a criterion that we will go by when assessing quality of your predictions for the test dataset.  You get these four points if your predictions for **test** dataset are better than those obtained from a fair coin flip (already shown in leaderboard and as examples of the file format for predictions upload) by at least 10% on **all** four metrics shown in the leaderboard (accuracy, sensitivity, specificity and precision).  But then predictions by the coin flip should not be very difficult to improve upon.  

# Extra 5 points: KNN model

Develop KNN classifiers for various values of $K$ and compare their performance to that from the rest of the models evaluated herein.  The challenges associated with this task include: a) handling the increase in computational demands of KNN models with the increase in the size of the data, and b) deciding on the choice of the distance that is suitable for a dataset with categorical attributes. 

# Extra 10 points: neural network model

Experiment with fitting neural network models of categorical outcome `response` for this data and evaluate their performance on different splits of the data into training and test. Compare model performance to that for the rest of classifiers developed above.


# An afterword on the computational demands of the final exam

Because during previous offerings of this course there were always several posts on piazza regarding how long it takes to fit various classifiers to the final exam dataset we have added this note here.

First of all, we most definitely do *not* expect you to *have* to buy capacity from AWS to complete this assignment. You certainly can if you want to, but this course is not about that and this dataset is really not *that* big to require it. Something reasonable/useful can be accomplished for this data with middle of the road hardware. For instance, knitting of the entire official solution for the final exam on 8Gb RAM machine with two i5-7200u cores takes about an hour using single-threaded R/Rstudio and this includes both extra points problems as well as various assessments of the performance of different models as function of data size and so on.

Second, your solution should not take hours and hours to compile. If it does, it could be that it is attempting to do too much, or something is implemented inefficiently, or just plain incorrectly - it is impossible for us to comment on this until we see the code when we grade it. In general, it is often very prudent to "start small" -- fit your model on a random subset of data small enough for the model fitting call to return immediately, check how model performance (both in terms of error and time it takes to compute) scales with the size of the data you are training it on (as you increase it in size, say, two-fold several times), for tuning start with very coarse grid of parameter values and given those results decide what it right for you, etc.

Lastly, making the decision about what is right for the problem at hand, how much is enough, etc. is inherent in this line of work. If you choose to conduct model tuning on a subset of the data - especially if you have some assessment of how the choice of tuning parameter and test error is affected by the size of training dataset - it could be a very wise choice.  If it is more efficient for you to knit each problem separately, by all means feel free to do that - just remember to submit each .Rmd and HTML file that comprises your entire solution. On that note, if you end up using any of the unorthodox setups for your calculations (e.g. AWS, parallel processing, multiple machines, etc. - none of which are essential for solving it correctly) please be sure that when we grade we have every relevant piece of code available - we won't be able to grade your work if we are not clear about how the results were obtained.

In the end, the final exam asks you to assess performance of several classification technologies on a new dataset, decide on which classifier is the best and use it to make predictions for the test data. It is very much up to you how exactly you want to go about it.  There could be many versions of correct and informative solution for that (as there could be just as many if not more that are completely wrong).

As always, best of luck - we are practically done here!
